spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data

### 2020-08-26 paper/spyrison-cook.Rmd
gifs to seprate htmls

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = TRUE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = FALSE,
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "dc8244c7e3f470fe3eee86f5c577c0dae21136c3")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

# step0, output
# Adjust centering to make axes and data side by side
oblique_frame(f_bas, data = f_dat, manip_var = f_mvar, lab = colnames(f_dat),
              color = f_clas, shape = f_clas) + 
  coord_fixed() + theme_void() + 
  theme(legend.position = "none")
```

@@

p1 <- oblique_frame(f_mtour[,, 1],  lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(1) phi = 0.78") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p2 <- oblique_frame(f_mtour[,, 5],  lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(2) phi = 0") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p3 <- oblique_frame(f_mtour[,, 12], lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(3) phi = 1.57") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p4 <- oblique_frame(f_mtour[,, 15], lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(4) phi = 0.78") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
grid.arrange(p1, p2, p3, p4, ncol = 4)
```

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = f_bas, data = f_dat, manip_var = f_mvar,
                   color = f_clas, shape = f_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("flea data, radial tour of aede2")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "flea_manualtour_mvar5.gif", 
                   gif_path = "./output")
}
```

@@

funcs <- rbind(
  c("construction", "create_manip_space", "forms the 3D space of rotation"),
  c("construction", "rotate_manip_space", "performs 3D rotation"),
  c("construction", "manual_tour", "generates sequence of 2D frames"),
  c("", "", ""),
  c("render", "array2df", "turn the tour path array into long form, for plotting"),
  c("render", "render_", "render long form as a ggplot2 objection for animation"),
  c("render", "render_plotly", "render the animation as a plotly object (default)"),
  c("render", "render_gganimate", "render the animation as a gganimate object"),
  c("", "", ""),
  c("animation", "play_tour_path", "composite function animating the specified tour path"),
  c("animation", "play_manual_tour", "composite function animating the specified manual tour"),
  c("", "", ""),
  c("specialty", "oblique_basis", "table of the rotated basis and manip space"),
  c("specialty", "oblique_frame", "display the reference axes of a given basis"),
  c("specialty", "view_manip_space", "illustrative display of any manip space")
)
colnames(funcs) <- c("Type", "Function", "Description")

@@

The function `oblique_frame` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

```{r eval=FALSE}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
```

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("jet cluster, radial tour of PC4")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_manualtour_pc4.gif", 
                   gif_path = "./output")
}

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), 
                                 ggtitle("jet cluster, radial tour of PC3")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_manualtour_pc3.gif", 
                   gif_path = "./output")
}
```

@@

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("DIS cluster, radial tour of PC6")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_manualtour_pc6.gif", 
                   gif_path = "./output")
}

@@

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), 
                                  ggtitle("jet cluster, radial tour of PC2")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_manualtour_pc2.gif", 
                   gif_path = "./output")
}
```

>>>>

---
title: > 
  spinifex: An R Package for Creating a Manual Tour of Low-dimensional Projections of Multivariate Data
author:
  - name: Nicholas Spyrison
    affiliation: Monash University
    address:
    - Faculty of Information Technology
    email:  nicholas.spyrison@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
abstract: >
  Dynamic low-dimensional linear projections of multivariate data collectively known as *tours* provide an important tool for exploring multivariate data and models. The R package \pkg{tourr} provides functions for several types of tours: grand, guided, little, local and frozen. Each of these can be viewed dynamically, or saved into a data object for animation. This paper describes a new package, \pkg{spinifex}, which provides a manual tour of multivariate data where the projection coefficient of a single variable is controlled. The variable is rotated fully into the projection, or completely out of the projection. The resulting sequence of projections can be displayed as an animation, with functions from either the \pkg{plotly} or \pkg{gganimate} packages. By varying the coefficient of a single variable, it is possible to explore the sensitivity of structure in the projection to that variable. This is particularly useful when used with a projection pursuit guided tour to simplify and understand the solution. The use of the manual tour is applied particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions.
keywords: exploratory data analysis, projection pursuit, high dimensional data, data visualization, statistical graphics, data science
preamble: >
  % preamble packages
link-citations: true
fig_caption: yes
toc: false
bibliography: spyrison-cook.bib
output: rticles::rjournal_article
  #bookdown::pdf_book ## alternative
---
```{r setup_paper, include=FALSE}
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = TRUE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = FALSE,
  cache.lazy = FALSE
)
if (F){ ## From reproducing, please use this dev ver:
  remotes::install_github("nspyrison/spinifex", 
                          ref = "176a5e9184e794109579e741cddc82931ceb0dfb")
}
library(tourr)
library(spinifex) ## Use dev version listed above!
library(ggplot2)
library(gridExtra)
library(dplyr)
```

@@

# step0, output
# Adjust centering to make axes and data side by side
view_frame(f_bas, data = f_dat, manip_var = f_mvar, lab = colnames(f_dat),
              color = f_clas, shape = f_clas) + 
  coord_fixed() + theme_void() + 
  theme(legend.position = "none")
```

@@

p1 <- view_frame(f_mtour[,, 1],  lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(1) phi = 0.78") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p2 <- view_frame(f_mtour[,, 5],  lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(2) phi = 0") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p3 <- view_frame(f_mtour[,, 12], lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(3) phi = 1.57") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
p4 <- view_frame(f_mtour[,, 15], lab = colnames(f_dat), manip_var = f_mvar) +
  ggtitle("(4) phi = 0.78") + coord_fixed() +
  theme_void() + theme(legend.position = "none")
grid.arrange(p1, p2, p3, p4, ncol = 4)
```

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = f_bas, data = f_dat, manip_var = f_mvar,
                   color = f_clas, shape = f_clas,
                   ggtheme = list(theme_void(), theme(legend.position = "none")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "flea_radialtour_mvar5.gif", 
                   gif_path = "./gifs")
}
```

@@

funcs <- rbind(
  c("construction", "create_manip_space", "forms the 3D space of rotation"),
  c("construction", "rotate_manip_space", "performs 3D rotation"),
  c("construction", "manual_tour", "generates sequence of 2D frames"),
  c("", "", ""),
  c("render", "array2df", "turn the tour path array into long form, for plotting"),
  c("render", "render_", "render long form as a ggplot2 objection for animation"),
  c("render", "render_plotly", "render the animation as a plotly object (default)"),
  c("render", "render_gganimate", "render the animation as a gganimate object"),
  c("", "", ""),
  c("animation", "play_tour_path", "composite function animating the specified tour path"),
  c("animation", "play_manual_tour", "composite function animating the specified manual tour"),
  c("", "", ""),
  c("specialty", "print_manip_space", "table of the rotated basis and manip space"),
  c("specialty", "view_frame", "display the reference axes of a given basis"),
  c("specialty", "view_manip_space", "illustrative display of any manip space")
)
colnames(funcs) <- c("Type", "Function", "Description")

@@

The function `view_frame` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

```{r eval=FALSE}
## View a basis and projected data 
view_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
```

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), theme(legend.position = "none")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_radialtour_pc4.gif", 
                   gif_path = "./gifs")
}

@@

## Saving gif for the supplementary material:
if (FALSE){
  play_manual_tour(basis = jet_bas, data = jet_dat, manip_var = jet_mvar,
                   color = jet_clas, shape = jet_clas,
                   ggtheme = list(theme_void(), theme(legend.position = "none")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "jetcluster_radialtour_pc3.gif", 
                   gif_path = "./gifs")
}
```

@@

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), theme(legend.position = "none")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_radialtour_pc6.gif", 
                   gif_path = "./gifs")
}

@@

## Saving gif for the supplementary material
if (FALSE){
  play_manual_tour(basis = DIS_bas, data = DIS_dat, manip_var = DIS_mvar,
                   color = DIS_clas, shape = DIS_clas,
                   ggtheme = list(theme_void(), theme(legend.position = "none")),
                   fps = 8, start_pause = .5, end_pause = 1,
                   render_type = render_gganimate,
                   gif_filename = "discluster_radialtour_pc2.gif", 
                   gif_path = "./gifs")
}
```

### 2020-08-26 paper/spyrison-cook.tex
gifs to seprate htmls

\caption{\label{tab:functionsTable}Summary of available functions.}
\centering
\begin{tabular}[t]{lll}
\toprule
Type & Function & Description\\
\midrule
construction & create\_manip\_space & forms the 3D space of rotation\\
construction & rotate\_manip\_space & performs 3D rotation\\
construction & manual\_tour & generates sequence of 2D frames\\
 &  & \\
render & array2df & turn the tour path array into long form, for plotting\\
render & render\_ & render long form as a ggplot2 objection for animation\\
render & render\_plotly & render the animation as a plotly object (default)\\
render & render\_gganimate & render the animation as a gganimate object\\
 &  & \\
animation & play\_tour\_path & composite function animating the specified tour path\\
animation & play\_manual\_tour & composite function animating the specified manual tour\\
 &  & \\
specialty & oblique\_basis & table of the rotated basis and manip space\\
specialty & oblique\_frame & display the reference axes of a given basis\\
specialty & view\_manip\_space & illustrative display of any manip space\\
\bottomrule
\end{tabular}
\end{table}

@@

The function \texttt{oblique\_frame} can be used to show a projection of
the basis, or with the data overlaid. For example, the plots in Figures
\ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

\begin{Schunk}
\begin{Sinput}
## View a basis and projected data 
oblique_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
\end{Sinput}
\end{Schunk}

>>>>

\caption{\label{tab:functionsTable}Summary of available functions.}
\centering
\begin{tabular}[t]{lll}
\toprule
Type & Function & Description\\
\midrule
construction & create\_manip\_space & forms the 3D space of rotation\\
construction & rotate\_manip\_space & performs 3D rotation\\
construction & manual\_tour & generates sequence of 2D frames\\
 &  & \\
render & array2df & turn the tour path array into long form, for plotting\\
render & render\_ & render long form as a ggplot2 objection for animation\\
render & render\_plotly & render the animation as a plotly object (default)\\
render & render\_gganimate & render the animation as a gganimate object\\
 &  & \\
animation & play\_tour\_path & composite function animating the specified tour path\\
animation & play\_manual\_tour & composite function animating the specified manual tour\\
 &  & \\
specialty & print\_manip\_space & table of the rotated basis and manip space\\
specialty & view\_frame & display the reference axes of a given basis\\
specialty & view\_manip\_space & illustrative display of any manip space\\
\bottomrule
\end{tabular}
\end{table}

@@

The function \texttt{view\_frame} can be used to show a projection of
the basis, or with the data overlaid. For example, the plots in Figures
\ref{fig:step0} and \ref{fig:step3} were made with code similar to this:

\begin{Schunk}
\begin{Sinput}
## View a basis and projected data 
view_frame(basis = f_basis, 
              data = f_data,
              color = f_clas,
              shape = f_clas)
\end{Sinput}
\end{Schunk}
